GLOBALIZER: A NOVEL SUPERCOMPUTER SOFTWARE SYSTEM FOR SOLVING TIME-CONSUMING GLOBAL OPTIMIZATION PROBLEMS

### 2016-08-04 2016/GOM/paper.tex
Update experiments section of GOM paper

\theoremstyle{definition}
\newtheorem{definition}{Definition}

@@

  x_{k+1}=\dfrac{x_t+x_{t-1}}{2}-sign(z_{t}-z_{t-1})\dfrac{1}{2r}\left[\dfrac{|z_{t}-z_{t-1}|}{\mu}\right]^N,\quad 1<t<k+1 \\

@@

\subsection{Parallel Computations for Systems with Shared Memory}
\label{subsec:sharedpar}
Modern supercomputer computational systems consist of plenty of computer nodes,
which include several multicore processors. At that, random access memory is
shared for the computing nodes --- the content of any memory element can be read (written)
by any computer cores available at any arbitrary moment. In most cases, shared memory is
uniform --- the time characteristics for accessing memory are the same for all computer
cores and for all memory elements.
\par
The following speculations can form the basis for selecting parallel computation
organization methods. As was mentioned above when describing the MAGS algorithm,
the characteristics \(R(i),1\leqslant i\leqslant k+1\), calculated according to (\ref{step3})
can be interpreted as some measures of the importance of intervals with respect to the location of
the global minimum point. Following this understanding, each successive global search
iteration is executed within the most important interval with the maximum value of the
characteristic. So far, in this case it is obvious how to select the other intervals for
organizing simultaneous computations of the minimized function values at several different
points within the search domain as well --- these should be the intervals with the next
magnitudes of characteristics.
\par
The computational scheme of the Parallel Multidimensional Algorithm of Global Search for
computer systems with shared memory (PMAGS) was developed based the approach considered
above and is practically identical to the MAGS scheme --- the differences consist just in the
following set of rules --- see (Strongin, R. G., Sergeyev, Y. D., 2000), (Strongin, R. G., K. A., Gergel, V. P., Grishagin, V. A., Barkalov, K. A.,  2013).
\par
\textit{Rule 4'}. To arrange the characteristics of the intervals obtained according to (\ref{step3}) in decreasing order
\begin{equation}
\label{step4par}
R(t_1)\geqslant R(t_2)\geqslant \dots \geqslant R(t_{k})\geqslant R(t_{k+1})
\end{equation}
and to select \(p\) intervals with the indices \(t_j,1\leqslant j\leqslant p\), with
the highest values of characteristics (\(p\) is the number of processors/cores used for the parallel computations).
\par
\textit{Rule 5'}. To execute new trials (computations of values for the minimized
function \(f(x)\)) at the points \(x_{k+j},1\leqslant j\leqslant p\), located in the
intervals with the highest characteristics from (\ref{step4par})
\begin{equation}
\label{step5par}
\begin{array}{cr}\\
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2}-sign(z_{t_j}-z_{t_j-1})\dfrac{1}{2r}\left[\dfrac{|z_{t_j}-z_{t_j-1}|}{\mu}\right]^N,\quad 1<t_j<k+1 \\
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2},\quad t_j=1,t_j=k+1
\end{array}
\end{equation}
\par
The stop condition for the algorithm (\ref{eq:stop}) should be checked for all
intervals from (\ref{step4par}), in which the scheduled trials are executed
\begin{equation}
  \label{eq:stop}
\rho_{t_j}<\varepsilon,1\leqslant j\leqslant p
\end{equation}
As before, if the stop condition is not fulfilled, the index \(k\) is
incremented by \(p\), and a new global search iteration is executed.
\par
The conditions for the developed parallel algorithm’s convergence and for the
non-redundancy of parallel computations have been examined in (Strongin, R. G., Sergeyev, Y. D., 2000).
Thus, in particular, when the conditions for convergence are satisfied, the parallel
computations are non-redundant as compared to the sequential method when using up to
\(2N\) processors/cores (where \(N\) is the dimensionality of the global optimization problem being solved).

@@

\subsection{Parallel Computations Using Intel Xeon Phi processors}
The results of the numerical experiments with GPMAGS on an Intel Xeon Phi are provided in Table 1.
The computations were performed using the Simple and Hard function classes with the dimensions equal to 4 and 5. The parameters  at  and  at  were used. When using the GPMAGS method, the parameter  was selected for the Simple class, and  was selected for the Hard class.
\par
In the first series of experiments, serial computations using MAGS were executed. The average number of iterations performed by the method for solving a series of problems for each of these classes is shown in row I. The symbol “>” reflects the situation where not all problems of a given class were solved by a given method. It means that the algorithm was stopped once the maximum allowable number of iterations Kmax was achieved. In this case, the Kmax value was used for calculating the average number of iterations corresponding to the lower estimate of this average value. The number of unsolved problems is specified in brackets.
\par
In the second series of experiments, parallel computations were executed on a CPU. The relative “speedup” in iterations achieved is shown in row II; the speedup of parallel computations was measured in relation to the sequential computations (p=1).

\subsection{Parallel Computations Using General Purpose Graphic Processors}

>>>>

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\DeclareMathOperator{\sign}{sign}

@@

  x_{k+1}=\dfrac{x_t+x_{t-1}}{2}-\sign(z_{t}-z_{t-1})\dfrac{1}{2r}\left[\dfrac{|z_{t}-z_{t-1}|}{\mu}\right]^N,\quad 1<t<k+1 \\

@@

\subsection{Parallel Computations for Systems with Shared Memory}
\label{subsec:sharedpar}
Modern supercomputer computational systems consist of plenty of computer nodes,
which include several multicore processors. At that, random access memory is
shared for the computing nodes --- the content of any memory element can be read (written)
by any computer cores available at any arbitrary moment. In most cases, shared memory is
uniform --- the time characteristics for accessing memory are the same for all computer
cores and for all memory elements.
\par
The following speculations can form the basis for selecting parallel computation
organization methods. As was mentioned above when describing the MAGS algorithm,
the characteristics \(R(i),1\leqslant i\leqslant k+1\), calculated according to (\ref{step3})
can be interpreted as some measures of the importance of intervals with respect to the location of
the global minimum point. Following this understanding, each successive global search
iteration is executed within the most important interval with the maximum value of the
characteristic. So far, in this case it is obvious how to select the other intervals for
organizing simultaneous computations of the minimized function values at several different
points within the search domain as well --- these should be the intervals with the next
magnitudes of characteristics.
\par
The computational scheme of the Parallel Multidimensional Algorithm of Global Search for
computer systems with shared memory (PMAGS) was developed based the approach considered
above and is practically identical to the MAGS scheme --- the differences consist just in the
following set of rules --- see (Strongin, R. G., Sergeyev, Y. D., 2000), (Strongin, R. G., K. A., Gergel, V. P., Grishagin, V. A., Barkalov, K. A.,  2013).
\par
\textit{Rule 4'}. To arrange the characteristics of the intervals obtained according to (\ref{step3}) in decreasing order
\begin{equation}
\label{step4par}
R(t_1)\geqslant R(t_2)\geqslant \dots \geqslant R(t_{k})\geqslant R(t_{k+1})
\end{equation}
and to select \(p\) intervals with the indices \(t_j,1\leqslant j\leqslant p\), with
the highest values of characteristics (\(p\) is the number of processors/cores used for the parallel computations).
\par
\textit{Rule 5'}. To execute new trials (computations of values for the minimized
function \(f(x)\)) at the points \(x_{k+j},1\leqslant j\leqslant p\), located in the
intervals with the highest characteristics from (\ref{step4par})
\begin{equation}
\label{step5par}
\begin{array}{cr}\\
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2}-\sign(z_{t_j}-z_{t_j-1})\dfrac{1}{2r}\left[\dfrac{|z_{t_j}-z_{t_j-1}|}{\mu}\right]^N,\quad 1<t_j<k+1 \\
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2},\quad t_j=1,t_j=k+1
\end{array}
\end{equation}
\par
The stop condition for the algorithm (\ref{eq:stop}) should be checked for all
intervals from (\ref{step4par}), in which the scheduled trials are executed
\begin{equation}
  \label{eq:stop}
\rho_{t_j}<\varepsilon,1\leqslant j\leqslant p
\end{equation}
As before, if the stop condition is not fulfilled, the index \(k\) is
incremented by \(p\), and a new global search iteration is executed.
\par
The conditions for the developed parallel algorithm’s convergence and for the
non-redundancy of parallel computations have been examined in (Strongin, R. G., Sergeyev, Y. D., 2000).
Thus, in particular, when the conditions for convergence are satisfied, the parallel
computations are non-redundant as compared to the sequential method when using up to
\(2N\) processors/cores (where \(N\) is the dimensionality of the global optimization problem being solved).

@@

\subsection{Parallel Computations Using Intel Xeon Phi processors}
The results of the numerical experiments with GPMAGS on an Intel Xeon Phi are provided in Table 1.
The computations were performed using the Simple and Hard function classes with the
dimensions equal to 4 and 5. The parameters \(\Delta=10^{-6}\) at \(N=4\) and \(\Delta=10^{-6}\) at \(N=5\)
were used. When using the GPMAGS method, the parameter \(r=4.5\) was selected for the
Simple class, and \(r=5.6\) was selected for the Hard class.
\par
In the first series of experiments, serial computations using MAGS were executed.
The average number of iterations performed by the method for solving a series of
problems for each of these classes is shown in row I. The symbol ``\(>\)'' reflects the
situation where not all problems of a given class were solved by a given method.
It means that the algorithm was stopped once the maximum allowable number of
iterations \(K_{max}\) was achieved. In this case, the \(K_{max}\) value was used for
calculating the average number of iterations corresponding to the lower estimate of
this average value. The number of unsolved problems is specified in brackets.
\par
In the second series of experiments, parallel computations were executed on a CPU.
The relative ``speedup'' in iterations achieved is shown in row II; the speedup of
parallel computations was measured in relation to the sequential computations (\(p=1\)).

\par
The final series of experiments was executed using a Xeon Phi. The results of these
computations are shown in row III; in this case, the speedup factor is calculated in
relation to the PMAGS results on a CPU using eight cores (\(p=8\)). It can be noted that
relatively high speedup was achieved. For example, solving a problem from the 5-Hard class
required on average only 633 parallel iterations on the Xeon Phi, whereas the number of
iterations was more than 37 thousand when using all of the CPU’s computing cores.

\subsection{Parallel Computations Using General Purpose Graphic Processors}
In the first series of experiments, only one graphic accelerator was used for solving
100 6-diminsional problems. The method’s reliability parameter \(r=4.5\) was selected.
\par
The average problem solving time using a GPU was 10.78 sec. whereas the averaged problem
solving time using all of the node’s 16 CPU cores was 53.8 sec.; an almost 5-fold improvement was observed.
\par
A larger scale computational experiment was carried out for the problems with
dimensions of \(N=8\) and \(N=10\): a total of 12 nodes within a cluster with 36
graphic accelerators (three accelerators per node) were employed, therefore, a total
of 96768 CUDA-cores were employed.
\par
The average time to solve the 8-dimensional problem was 405.6 seconds, 5.9 times
faster than the CPU version of the algorithm. The average time to solve the
10-dimensional problem was 2,055.8 sec. The CPU was not used to solve the 10-dimensional
problem because of the high degree of complexity in the computations.
